---
title: "Stats and plots"
author: "Alicia Chen"
date: "2024-05-11"
output:
  pdf_document: default
---

```{r setup}
library(here)
library(tidyverse)
library(lme4)
library(lmerTest)
library(tidyboot)
library(broom)
library(emmeans)
library(car)
library(jsonlite)

theme_set(theme_classic(base_size = 20))
```

# Load data

```{r}
all_calculations <- read.csv("outputs/all_calculations.csv") %>%
    mutate(
    paired_comm_score = (own_score + comm_score) / 2,
    learn_score_norm = (learn_score - min(learn_score, na.rm = TRUE)) / (max(learn_score, na.rm = TRUE) - min(learn_score, na.rm = TRUE)) ,
    alignment_norm = (alignment - min(alignment, na.rm = TRUE)) / (max(alignment, na.rm = TRUE) - min(alignment, na.rm = TRUE))
  )

df_within <- read.csv("outputs/metrics/within_clust_syst.csv") # %>% filter(min_cluster_size == 3)
df_between <- read.csv("outputs/metrics/btwn_clust_syst.csv") # %>% filter(min_cluster_size == 3)
df_systematicity <- read.csv("outputs/metrics/systematicity.csv")

# Merging data for each participant, add comm_score from all_calculations
df_within <- merge(df_within, all_calculations[, c("speaker", "game", "comm_score")], by = "speaker")
df_between <- merge(df_between, all_calculations[, c("speaker", "game", "comm_score")], by = "speaker")
df_systematicity <- merge(df_systematicity, all_calculations[, c("speaker", "game", "comm_score")], by = "speaker")
df_systematicity <- merge(df_systematicity, df_between[, c("speaker", "min_cluster_size", "n_clusters")], by = "speaker")

# Filter between cluster systematicity to only include participants with more than 1 cluster
df_between <- df_between[df_between$n_clusters > 1, ]

# Communication score per trial
d.comm <- read_csv(here('outputs/comm.zip'))

# Embeddings
d.1d <- read.csv(here('outputs/embedding_1d.csv')) %>%
  mutate(game = ifelse(speaker == 'init', 'zzz_init', game))

d.2d <- read.csv(here("outputs/embedding_2d.csv"))
```

# How well did participants do during the learning and communication phases? 

Does learning performance predict communication performance?

```{r}
ggplot(all_calculations, aes(x = learn_score_norm, y = comm_score)) +
  geom_point(size = 4,
             alpha = 0.8,
             stroke = 0) +
  geom_smooth(
    method = 'lm',
    se = TRUE,
    fill = "lightblue",
    color = "black"
  ) +
  labs(x = "learn score", y = "communication score") +
  geom_hline(yintercept = 0.5,
             color = "red",
             linetype = "dashed") +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()) +
  ylim(0.35, 1)
```

```{r}
lm(comm_score ~ 1 + learn_score_norm, data = all_calculations) %>% 
  summary()
```
## Performance across rounds

Do participants successfully communicate with each other? Does communication performance improve across rounds? 

```{r}
d.learn.comm <- d.comm %>%
  group_by(round) %>%
  tidyboot_mean(score, na.rm = T)

ggplot(d.learn.comm, aes(x = round, y = empirical_stat)) +
  geom_line() +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
  geom_hline(yintercept = 0.5,
             color = "red",
             linetype = "dashed") +
  labs( y = "communication score") +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()) +
  ylim(0.35, 1)

ggsave(here("figs/comm_improvement.pdf"),
       width = 4,
       height = 4.5)
```
Increase in performance over rounds

```{r}
comm.score.mod <- lmer(score ~ 1 + round + (1 |
                                                     gameid) + (1 | speakerid),
                              data = d.comm) 

comm.score.mod %>% summary()
```

Is performance above chance?

```{r}
emm <- emmeans(comm.score.mod, specs = "1")
summary(emm)

contrast_result <- contrast(emm, method = list("compare to 0.5" = c(1)), offset = -0.5)
summary(contrast_result)
```
# Representing and visualizing signals

## Generate 2d embedding figure

```{r}
# Make color mapping vector
col <- unique(d.2d$referent)
names(col) <- col

f = ggplot(d.2d, aes(x = mds_1, y = mds_2)) +
  geom_point(
    aes(color = referent),
    size = 2.7,
    alpha = 0.5,
    stroke = 0
  ) +
  geom_point(
    data = d.2d %>% filter(speaker == 'init'),
    aes(fill = referent),
    size = 4,
    shape = 22,
    stroke = 1.5
  ) +
  scale_color_manual(values = col) +
  scale_fill_manual(values = col) +
  theme(legend.position = "none") +
  coord_fixed() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
lims(x = c(-7, 6), y = c(-5, 5))

f

ggsave(here("figs/mds_all_init.pdf"),
       f,
       width = 4,
       height = 6)
```

## Generate 1d embedding figure

```{r, fig.width=18, fig.height=10}
# Make color mapping vector
col <- unique(d.1d$referent)
names(col) <- col

# Load in performance data and add to d.1d
d.1d <- merge(d.1d, all_calculations %>% select(c("game", "speaker", "comm_score")), all.x = T, by = c('game', 'speaker'))


data <- d.1d
data$speaker <- as.factor(data$speaker)

# Calculate position based on comm_score within each game and speaker
position_data <- data %>%
  group_by(game, speaker) %>%  # Group by both game and speaker
  summarise(comm_score = mean(comm_score, na.rm = TRUE), .groups = 'drop') %>%
  arrange(game, comm_score) %>%
  mutate(position = if_else(row_number() %% 2 == 1, -0.1, 0.1)) %>%  # Assign -1 or 1 based on sorted order
  mutate(position = ifelse(speaker == 'init', 0, position))  # Adjust position for "init" speaker

# Join calculated positions back to the original dataset
data <- data %>%
  left_join(position_data %>% select(-comm_score), by = c("game", "speaker"))

# Adjust game_order to include a specific game for "init" at the bottom
game_order <- data %>%
  group_by(game) %>%
  summarise(avg_comm_score = mean(comm_score, na.rm = TRUE)) %>%
  arrange(desc(avg_comm_score)) %>%
  mutate(game = if_else(game == "init", "zzz_init", game)) %>%
  arrange(desc(avg_comm_score), .by_group = TRUE) %>%
  pull(game)

# Order game factor based on average communication score
data$game <- factor(data$game, levels = game_order)

# Plotting
ggplot(data, aes(x=mds_1, y=position, color=referent)) +
   geom_point(data = subset(data, game == "zzz_init"), aes(fill = referent),
              shape = 22,
              stroke = 1.5,
              size = 7) +
   geom_point(data = subset(data, game != "zzz_init"), aes(color = referent),
              size = 4,
              stroke = 0,
              alpha = 0.8) +
  scale_fill_manual(values = col) +
  scale_color_manual(values = col) +
  scale_y_continuous(name="", breaks=NULL, limits = c(-0.2, 0.2)) +  # Remove y-axis labels
  labs(x="MDS_1", color="Referent") +
  facet_wrap(~ game, ncol = 1) +
  theme(legend.position="none",
        strip.text = element_blank(),  # Remove the game labels from the facet strips
        strip.background = element_blank())

ggsave(here("figs/1d_embeddings.pdf"), width = 18, height = 10)
```

# What properties led to successful communication?

## Partner alignment

```{r}
ggplot(
  all_calculations %>% distinct(alignment_norm, .keep_all = T),
  aes(x = alignment_norm, y = paired_comm_score)
) +
  geom_point(size = 5,
             alpha = 0.8,
             stroke = 0) +
  geom_smooth(
    method = 'lm',
    se = TRUE,
    fill = "lightblue",
    color = "black"
  ) +
  geom_hline(yintercept = 0.5,
             color = "red",
               linetype = "dashed") +
    theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()) +
    ylim(0.35, 1)

ggsave(here("figs/alignment.pdf"),
       width = 5,
       height = 4.5)
```
Correlate partner alignment with communication

```{r}
all_calculations %>% distinct(alignment_norm, paired_comm_score) %>%
  cor.test(~ alignment_norm + paired_comm_score, data = .)
```
How much is this explained by how much participants are able to remember signal-color pairings in the first place? 

```{r}
lm(comm_score ~ 1 + alignment_norm + learn_score, data = all_calculations) %>% 
  summary()
```
## Quantitative analysis of strategies 

### Discreteness

#### Hopkins statistic

Calculate mean and SD of hopkins
```{r}
mean(all_calculations$hopkins)
sd(all_calculations$hopkins)
```

Plot Hopkins statistic vs. communication score

```{r}
ggplot(all_calculations, aes(x = hopkins, y = comm_score, color = systematicity)) +
  geom_point(size = 5) +
  geom_smooth(method = "lm", se = T, color = "black") +
    geom_hline(yintercept = 0.5,
             color = "red",
               linetype = "dashed") +
  labs(y = "communication score", color = "dcor") +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "bottom")

ggsave(here("figs/hopkins_comm.pdf"),
       width = 7,
       height = 6)
```

Correlate Hopkins statistic with communication score

```{r}
cor.test(all_calculations$hopkins, all_calculations$comm_score)
```

#### Number of clusters

Number of clusters vs. communication score for `min_cluster_size = 2` in HDBSCAN

```{r}
# Look at number of clusters vs systematicity
ggplot(df_systematicity %>% filter(min_cluster_size == 2), aes(x = n_clusters, y = comm_score, color = dcor)) +
  geom_point(size = 5, alpha = 0.8) +
  geom_smooth(method = "lm", se = T, color = "black") +
  labs(x = "number of clusters", y = "communication score") +
  scale_x_continuous(breaks = seq(1, 7, 1)) +
    geom_hline(yintercept = 0.5,
             color = "red",
               linetype = "dashed") +
  theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank())

ggsave(here("figs/n_clusters_comm.pdf"),
       width = 7,
       height = 4.5)
```

```{r}
cor.test(all_calculations$n_clusters, all_calculations$comm_score)
```
Count the number of participants that had 5 clusters

```{r}
df_systematicity %>% filter(min_cluster_size == 2, n_clusters == 5) %>% nrow()
```

### Systematicity


#### Global systematicity

P values

```{r}
ggplot(df_systematicity %>% filter(min_cluster_size == 2), aes(x = dcor, y = p, color = comm_score)) +
  geom_point(size = 5, alpha = 0.8) +
  geom_hline(yintercept = 0.05, color = "red", linetype = "dashed") +
  labs(x = "systematicity", y = "p value", title = "systematicity - p values by communication score") +
  scale_color_viridis_c(option = "plasma", name = "comm score") +
    theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="bottom")

ggsave(here("figs/sys_p.pdf"),
       width = 7,
       height = 6)
```
How many participants had p-values less than 0.05? 

```{r}
df_systematicity %>% filter(min_cluster_size == 2, p < 0.05) %>% nrow()
```

Display systematicity for init signals
```{r}
init_sys <- read.csv("outputs/metrics/systematicity.csv") %>% filter(speaker == "init")
init_sys
```

Correlate systematicity with communication score

```{r}
cor.test(all_calculations$systematicity, all_calculations$comm_score)
```


Systematicity vs. communication score, colored by number of clusters

```{r}
ggplot(df_systematicity %>% filter(min_cluster_size == 2), aes(x = dcor, y = comm_score, color = as.factor(n_clusters))) +
  geom_point(size = 5, alpha = 0.8) +  # Set size and transparency of points
  geom_smooth(method = "lm", se = T, color = "black") +
    geom_hline(yintercept = 0.5, color = "red", linetype = "dashed") +
  scale_color_viridis_d(name = "n_clusters") +  # Viridis discrete color scale for 'n_clusters'
  labs(x = "systematicity", y = "communication score", title = "communication score vs systematicity") +
    theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position="bottom")
```

Systematicity vs. commnunication score, colored by hopkins

```{r}
ggplot(all_calculations, aes(x = systematicity, y = comm_score, color = hopkins)) +
  geom_point(size = 5, alpha = 0.8) + 
  geom_smooth(method = "lm", se = T, color = "black") +
    geom_hline(yintercept = 0.5, color = "red", linetype = "dashed") +
  scale_color_viridis_c(name = "hop") +  # Viridis discrete color scale for 'n_clusters'
  labs(x = "systematicity", y = "communication score") +
    theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        )

ggsave(here("figs/sys_comm_hopkins.pdf"),
       width = 7,
       height = 4.5)
```
#### Systematicity in the presence of discreteness

```{r}
df_sys_with_between <- merge(
  df_systematicity %>% filter(min_cluster_size == 2),
  df_between %>% filter(min_cluster_size == 2) %>% select(c(speaker, dcor)) %>% rename(dcor_between = dcor),
  by = "speaker"
) %>%
  filter(n_clusters >= 3)
```

How many participants had statistically significant amounts of between-cluster systematicity?

```{r}
df_between %>% filter(min_cluster_size == 2) %>% nrow()
df_between %>% filter(min_cluster_size == 2, p > 0.05) %>% nrow()
```


How many clusters had statistically significant amounts of within-cluster systematicity?

```{r}
df_within %>% filter(min_cluster_size == 2) %>% nrow()
df_within %>% filter(min_cluster_size == 2, p < 0.05, n_signals < 40) %>% nrow()
```

Group by participants, check how many clusters per participant had significant within-cluster systematicity

```{r}
df_within %>% filter(min_cluster_size == 2) %>% group_by(speaker) %>% filter(p < 0.05) %>% summarise(n_clusters = n())
```



Is between cluster systematicity correlated with general systematicity?

```{r}
cor.test(df_sys_with_between$dcor_between, df_sys_with_between$dcor)
```
Is between-cluster systematicity correlated with communication score? 

```{r}
cor.test(df_sys_with_between$dcor_between, df_sys_with_between$comm_score)
```

For robustness: Repeat the analyses in this section, looping through different values of `min_cluster_size`

```{r}
for (min_cluster_size_ in unique(df_systematicity$min_cluster_size)) {
  df_sys_with_between <- merge(
    df_systematicity %>% filter(min_cluster_size == min_cluster_size_),
    df_between %>% filter(min_cluster_size == min_cluster_size_) %>% select(c(speaker, dcor)) %>% rename(dcor_between = dcor),
    by = "speaker"
  ) %>%
    filter(n_clusters >= 3)
  
  print(paste("min_cluster_size:", min_cluster_size_))
  print(cor.test(df_sys_with_between$dcor_between, df_sys_with_between$dcor))
  print(cor.test(df_sys_with_between$dcor_between, df_sys_with_between$comm_score))
}
```


## Big regression: Systematicity and alignment predict communication score

Scale all predictors

```{r}
all_calculations_scaled <- all_calculations %>%
  mutate(across(!c(game, speaker), ~ scale(.) %>% as.vector()))
```

```{r}
mod <- lm(comm_score ~ learn_score + alignment + systematicity + hopkins + n_clusters, data = all_calculations_scaled)
summary(mod)
```

Plot coefficients

```{r}
tidy_model <- tidy(mod, conf.int = TRUE) %>% filter(term != "(Intercept)")
tidy_model$term <- factor(tidy_model$term, levels = rev(c("learn_score", "alignment", "systematicity", "n_clusters", "hopkins")))

ggplot(tidy_model, aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high)) +
  geom_pointrange(fatten = 5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs( x = "standardized coefficient estimate",
       y = "predictor")

ggsave(here("figs/coefficients.pdf"),
       width = 4.5,
       height = 4.5)
```


## Supplement

Stress plot with MDS dimensions

```{r}
stresses <- read_json(here("test/stresses.json")) %>%
  as.data.frame() %>%
  setNames(1:5) %>%
  pivot_longer(cols = everything(), names_to = "components", values_to = "stresses") %>%
  mutate(components = as.numeric(components))
```

```{r}
ggplot(data = stresses, aes(x = components, y = stresses)) +
  geom_point() +
  geom_line() +
  labs(x = "components", y = "stress") +
    theme(axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank())

ggsave(here("figs/mds_scree.pdf"),
       width = 6,
       height = 4)
```




